Assumptions of Physics blueprint: information granularity as a common foundation for geometry, measures and probability

### 2021-05-19 Overviews/Informational granularity/InformationGranularity.tex
InfGran: version pushed on arxiv

	Transitivity is unclear. Let $\stmt_1 \equ \stmt_2 \equ \stmt_3$. Then we can write $\stmt_1 \OR \stmt[v]_1 \eqgran \stmt_2 \OR \stmt[v]_2$ and $\stmt_2 \OR \stmt[w]_2 \eqgran \stmt_3 \OR \stmt[w]_3$. If the infinitesimal where incompatible we would have $\stmt_1 \OR \stmt[v]_1 \OR \stmt[w]_2 \eqgran \stmt_2 \OR \stmt[v]_2 \OR \stmt[w]_2 \eqgran \stmt_3 \OR \stmt[w]_3 \OR \stmt[v]_2$. If the infinitesimals were comparable, by proposition \ref{infinitesimal_disjunction} we would have that $\stmt[v]_1 \OR \stmt[w]_2$ and $\stmt[w]_3 \OR \stmt[v]_2$ are infinitesimals and therefore $\stmt_1 \equ \stmt_3$.

	If we assume a uniform domain, we would have totality and we may also be able to show that incompatible infinitesimals can always be found. It may be that totality is not needed for proposition \ref{infinitesimal_disjunction}. It may be that we need to amend the definition of infinitesimals, such that the disjunction of two infinitesimals is an infinitesimals. We could also force the transitivity, even if the disjunction of two infinitesimals is not an infinitesimals. This would be like saying that, yes, in principles the assertions correspond to different granularity levels, but we are not to distinguish between them because the different is effectively zero. Understanding which options are best requires more work.
\end{remark}

@@

\begin{desid}
	Let $\stmt_1, \stmt_2 \in \tdomain$ such that $\stmt_1 \equ \stmt_2$ and $\stmt_1 \finer \stmt_2$. Then for all $\stmt_1 \finer \stmt \finer \stmt_2$ we have $\stmt_1 \equ \stmt \equ \stmt_2$.
\end{desid}

@@

\begin{desid}
	Let $\stmt_1,\stmt,\stmt[v]_1\in\tdomain$ such that $\stmt_1\finer\stmt\finer\stmt_1\OR\stmt[v]_1$, and suppose $\stmt[v]_1$ is infinitesimal with respect to $\stmt[u]$. Then $\stmt\equ\stmt_1$. 
\end{desid}

@@

Question: is the map $\tdomain / \eqgran \to \tdomain / \equ$ defined by collapsing equivalence classes well-defined and monotone?

@@

The established results in the literature of comparable probability require a specific axiom, called monotone continuity, that essentially make sure that sequences converge. This is necessary because their goal is to recover a probability measure that is countably additive.

@@

Note that, in general, monotone continuity is not satisfied. See examples in \ref{lack_of_monotone_continuity}. In terms of measures, the issue is that we may have that the disjoint disjuntion of countably many infinitesimals is not necessarily an infinitesimal, which means countably meany sets of measure zero would not sum up to a set of measure zero.

@@

The paper \cite{villegas} by Villegas gives us some conditions under which we can recover a probability measure given a preorder. Here we state the conditions required in terms of $\lequ$ and $\tdomain_{\stmt[u]}$, which we take it to be the set of statements narrower than $\stmt[u]$. We start from the axioms of pre-order:

@@

Next is the axiom of monotony. Let $\stmt_1,\stmt_2,\stmt[t]_1,\stmt[t]_2 \in \tdomain_{\stmt[u]}$ with $\stmt[t]_1 \ncomp \stmt[t]_2$. Then from $\stmt_1 \lequ \stmt[t]_1$ and $\stmt_2 \lequ \stmt[t]_2$ if follows that $\stmt_1\OR \stmt_2 \lequ \stmt[t]_1\OR \stmt[t]_2$. Moreover, if $\lequ$ is replace by $\ltu$ in one of the first two equality, then the last one holds with $\ltu$.

@@

Take $\mathbb{N}$ as the unit. For any $i$, we have $D_{i,j} \finer D_{i,l} \finer D_{i,j}$. Therefore $D_{i,j} \eqgran D_{i,l}$ for any $j$ and $l$. We have $1 = \mu_{\mathbb{N}}(\mathbb{N}) = \mu_{\mathbb{N}}(\bigOR_{j=1}^i D_{i,j}) = \sum_{j=1}^i \mu_{\mathbb{N}}( D_{i,j}) = \sum_{j=1}^i \mu_{\mathbb{N}}( D_{i,0}) = i \mu_{\mathbb{N}}( D_{i,0}) $. Therefore $\mu_{\mathbb{N}}( D_{i,j}) = 1/i$ for any $i$ and $j$.

@@

Consider the reals with the standard measure. A set of finitely many points points has measure zero. The rationals are countable and have measure zero. The Cantor set is uncountable and have measure zero. Therefore sets of real numbers with all possible cardinality can have measure zero. On the other hand, any finite open interval has uncountably many points and non-zero measure. Therefore two sets with the same cardinality can have different measure.

@@

Consider the following two sequences of subsets of $\mathbb{N}$. Let $\stmt_n = \{1,2,3,\ldots,n\}$, and let $\stmt[t]_n = \{2,4,6,\ldots,2n\}$. If we are treating $\mathbb{N}$ as a uniform domain, we should have $\stmt_n\eqgran\stmt[t]_n$ for all $n$. However, the limit $\bigOR_n\stmt_n = \mathbb{N}$, while $\bigOR_n\stmt[t]_n = \{2,4,6,\ldots\} = 2\mathbb{N} \subset \mathbb{N}$. These are not equigranular limits despite the sequences being equigranular.

@@

The problem is that infinite sets disrupt the relationship between narrowness and fineness, such that a saturated chain in narrowness is not necessarily a saturated chain in fineness. For example, if we have a countable set of possibilities (e.g. the natural numbers), a saturated chain in narrowness can start from the empty set, start by adding possibilities one be one until we end with the full set. The set with all possibilities except one, for example, cannot be added to the chain, but still represents a level of description that is in between all the finite sets and the full infinite set. The only case where we can construct saturated chains is precisely when the possibilities are finite, and is therefore not interesting.

@@

\begin{defn}
A theoretical domain $\tdomain$ is said to have the \textbf{saturated chains property} (SCP) if given any chain $S\subseteq\tdomain$ totally ordered by narrowness ($\narrower$), the following holds. There exists a maximum chain $S'\supseteq S$, and a saturated chain $C$ in the set of equigranularity equivalence classes, such that for all statements $s \in S'$, there exists an equivalence class $\overline{c}\in C$ with $s\in \overline{c}$.
\end{defn}

@@

\begin{prop}\label{chainimpliesuniform}
Let $\tdomain$ be a theoretical domain with the saturated chains property, and suppose all possibilities are comparable. Then they must be equigranular.
\end{prop}
\begin{proof}
Suppose all possibilities are comparable. Let $p,q$ be possibilities. Consider the two chains $\{\impossibility, p\}$ and $\{\impossibility, q\}$. By the SCP, these may be extended to saturated chains in narrowness, with corresponding saturated chains in fineness equivalence classes. Then since possibilities are successors of the impossibility, the fineness equivalence classes represent by $p$ and $q$ must both be an immediate successor of the equivalence class representing the impossibility. It follows that $p\finer q$, and similarly $q\finer p$, and so $p\eqgran q$. 
\end{proof}

\begin{prop}
Let $\tdomain$ be a theoretical domain with the saturated chains property. Then they must have finitely many possibilities.
\end{prop}
\begin{proof}
Let $\tdomain$ be a theoretical domain with the saturated chains property and such that the set $X$ of possibilities is infinite. Let $X' = \{x_i\}_{i=0}^{\infty} \subseteq X$ be a countable subset. Let $s_j = \bigOR_{j=0}^{i} x_i$ be the disjunction of the first $j$ possibilities. Let $\stmt = \bigOR_{i=0}^{\infty} x_i$ be the disjuction of all the possibilities in the sequence. The set $C = \{ \impossibility, \stmt \} \cup \{\stmt_j\}_{j=0}^{\infty}$ is a chain. In fact $\impossibility \snarrower \stmt_j \snarrower \stmt$ for all $j$ and $\stmt_i \narrower \stmt_j$ if $i \leq j$.

To show that $C$ is saturated, take a statement $\hat{\stmt} \narrower \stmt$ such that $\hat{\stmt} \notin C$. Let $A \subseteq X'$ be the set of possibilities compatible with $\hat{\stmt}$. If $A$ is infinite, since $\hat{\stmt} \nequiv \stmt$, there exists an $x_i \in X'$ such that $x_i \notin A$. Therefore $\stmt_{i} \nnarrower \hat{\stmt}$ since the first is compatible with a possibility that is not compatible with the second, and $\hat{\stmt} \nnarrower \stmt_{i}$ since $\hat{\stmt}$ is compatible with infinitely many statements and $\stmt_{i}$. If $A$ is finite, let the $x_i \in A$ for which $i$ is highest. Note that $A$ cannot contain all previous $x_j$ in the sequence, since $\hat{\stmt} \notin C$. Let $x_j$ the possibility with the highest $j \leq i$ that is not in $A$. We have $\hat{\stmt} \nnarrower \stmt_{j}$ and $\stmt_{i} \nnarrower \hat{\stmt}$ since each is compatible with a possibility incompatible with the other. Since we cannot add other statement, $C$ is saturated.

Now we show that, even though $C$ is saturated, it does not go through all levels of granularity. Let $x \in X'$ and consider $\stmt \AND \NOT x$. This is finer than $\stmt$ and is coarser than all other statements statements in $C$ since they are all compatible with finitely many possibilities. Therefore SCP does not hold. Therefore, if SCP holds, the domain cannot have infinitely many possibilities.
\end{proof}

>>>>

	Transitivity is unclear. Let $\stmt_1 \equ \stmt_2 \equ \stmt_3$. Then we can write $\stmt_1 \OR \stmt[v]_1 \eqgran \stmt_2 \OR \stmt[v]_2$ and $\stmt_2 \OR \stmt[w]_2 \eqgran \stmt_3 \OR \stmt[w]_3$. If the infinitesimals were incompatible we would have $\stmt_1 \OR \stmt[v]_1 \OR \stmt[w]_2 \eqgran \stmt_2 \OR \stmt[v]_2 \OR \stmt[w]_2 \eqgran \stmt_3 \OR \stmt[w]_3 \OR \stmt[v]_2$. If the infinitesimals were comparable, by proposition \ref{infinitesimal_disjunction} we would have that $\stmt[v]_1 \OR \stmt[w]_2$ and $\stmt[w]_3 \OR \stmt[v]_2$ are infinitesimals and therefore $\stmt_1 \equ \stmt_3$.

	If we assume a uniform domain, we would have totality and we may also be able to show that incompatible infinitesimals can always be found. It may be that totality is not needed for proposition \ref{infinitesimal_disjunction}. It may be that we need to amend the definition of infinitesimals, such that the disjunction of two infinitesimals is an infinitesimal. We could also force the transitivity, even if the disjunction of two infinitesimals is not an infinitesimal. This would be like saying that, yes, in principle the assertions correspond to different granularity levels, but we are not to distinguish between them because the difference is effectively zero. Understanding which options are best requires more work.
\end{remark}

@@

\begin{desid}
	Let $\stmt_1, \stmt_2 \in \tdomain$ be such that $\stmt_1 \equ \stmt_2$ and $\stmt_1 \finer \stmt_2$. Then for all $\stmt_1 \finer \stmt \finer \stmt_2$ we have $\stmt_1 \equ \stmt \equ \stmt_2$.
\end{desid}

@@

\begin{desid}
	Let $\stmt_1,\stmt,\stmt[v]_1\in\tdomain$ be such that $\stmt_1\finer\stmt\finer\stmt_1\OR\stmt[v]_1$, and suppose $\stmt[v]_1$ is infinitesimal with respect to $\stmt[u]$. Then $\stmt\equ\stmt_1$. 
\end{desid}

@@

Question: is the map $\tdomain / \eqgran \to \tdomain / \equ$ defined by collapsing equivalence classes well-defined and monotonic?

@@

The established results in the literature of comparable probability require a specific axiom, called monotone continuity, that essentially makes sure that sequences converge. This is necessary because their goal is to recover a probability measure that is countably additive.

@@

Note that, in general, monotone continuity is not satisfied. See examples in \ref{lack_of_monotone_continuity}. In terms of measures, the issue is that we may have that the disjoint disjunction of countably many infinitesimals is not necessarily an infinitesimal, which means countably many sets of measure zero would not sum up to a set of measure zero.

@@

The paper \cite{villegas} by Villegas gives us some conditions under which we can recover a probability measure given a preorder. Here we state the conditions required in terms of $\lequ$ and $\tdomain_{\stmt[u]}$, which we take to be the set of statements narrower than $\stmt[u]$. We start from the axioms of pre-order:

@@

Next is the axiom of monotony. Let $\stmt_1,\stmt_2,\stmt[t]_1,\stmt[t]_2 \in \tdomain_{\stmt[u]}$ with $\stmt[t]_1 \ncomp \stmt[t]_2$. Then from $\stmt_1 \lequ \stmt[t]_1$ and $\stmt_2 \lequ \stmt[t]_2$ if follows that $\stmt_1\OR \stmt_2 \lequ \stmt[t]_1\OR \stmt[t]_2$. Moreover, if $\lequ$ is replaced by $\ltu$ in one of the first two inequalities, then the last one holds with $\ltu$.

@@

Take $\mathbb{N}$ as the unit. For any $i$, we have $D_{i,j} \finer D_{i,l} \finer D_{i,j}$. Therefore $D_{i,j} \eqgran D_{i,l}$ for any $j$ and $l$. We have $1 = \mu_{\mathbb{N}}(\mathbb{N}) = \mu_{\mathbb{N}}(\bigOR_{j=0}^{i-1} D_{i,j}) = \sum_{j=0}^{i-1} \mu_{\mathbb{N}}( D_{i,j}) = \sum_{j=0}^{i-1} \mu_{\mathbb{N}}( D_{i,0}) = i \mu_{\mathbb{N}}( D_{i,0}) $. Therefore $\mu_{\mathbb{N}}( D_{i,j}) = 1/i$ for any $i$ and $j$.

@@

Consider the reals with the standard measure. A set of finitely many points has measure zero. The rationals are countable and have measure zero. The Cantor set is uncountable and has measure zero. Therefore sets of real numbers with all possible cardinality can have measure zero. On the other hand, any finite open interval has uncountably many points and non-zero measure. Therefore two sets with the same cardinality can have different measure.

@@

Consider the following two sequences of subsets of $\mathbb{N}$. Let $\stmt_n = \{1,2,3,\ldots,n\}$, and let $\stmt[t]_n = \{2,4,6,\ldots,2n\}$. If we are treating $\mathbb{N}$ as a uniform domain, we should have $\stmt_n\eqgran\stmt[t]_n$ for all $n$. However, the limit $\bigOR_n\stmt_n = \mathbb{N}$, while $\bigOR_n\stmt[t]_n = \{2,4,6,\ldots\} \subset \mathbb{N}$. These are not equigranular limits despite the sequences being equigranular.

@@

The problem is that infinite sets disrupt the relationship between narrowness and fineness, such that a saturated chain in narrowness is not necessarily a saturated chain in fineness. For example, if we have a countable set of possibilities (e.g. the natural numbers), a saturated chain in narrowness can start from the empty set, start by adding possibilities one by one until we end with the full set. The set with all possibilities except one, for example, cannot be added to the chain, but still represents a level of description that is in between all the finite sets and the full infinite set. The only case where we can construct saturated chains is precisely when the possibilities are finite, and is therefore not interesting.

@@

\begin{defn}
A theoretical domain $\tdomain$ is said to have the \textbf{saturated chains property} (SCP) if given any chain $S\subseteq\tdomain$ totally ordered by narrowness ($\narrower$), the following holds. There exists a saturated chain $S'\supseteq S$, and a saturated chain $C$ in the set of equigranularity equivalence classes, such that for all statements $s \in S'$, there exists an equivalence class $\overline{c}\in C$ with $s\in \overline{c}$.
\end{defn}

@@

\begin{prop}\label{chainimpliesuniform}
Let $\tdomain$ be a theoretical domain with the saturated chains property, and suppose all possibilities are comparable. Then they must be equigranular.
\end{prop}
\begin{proof}
Suppose all possibilities are comparable. Let $p,q$ be possibilities. Consider the two chains $\{\impossibility, p\}$ and $\{\impossibility, q\}$. By the SCP, these may be extended to saturated chains in narrowness, with corresponding saturated chains in fineness equivalence classes. Then since possibilities are successors of the impossibility, the fineness equivalence classes represented by $p$ and $q$ must both be an immediate successor of the equivalence class representing the impossibility. It follows that $p\finer q$, and similarly $q\finer p$, and so $p\eqgran q$. 
\end{proof}

\begin{prop}
Let $\tdomain$ be a theoretical domain with the saturated chains property. Then it must have finitely many possibilities.
\end{prop}
\begin{proof}
Let $\tdomain$ be a theoretical domain with the saturated chains property and such that the set $X$ of possibilities is infinite. Let $X' = \{x_i\}_{i=0}^{\infty} \subseteq X$ be a countable subset. Let $s_j = \bigOR_{j=0}^{i} x_i$ be the disjunction of the first $j$ possibilities. Let $\stmt = \bigOR_{i=0}^{\infty} x_i$ be the disjuction of all the possibilities in the sequence. The set $C = \{ \impossibility, \stmt \} \cup \{\stmt_j\}_{j=0}^{\infty}$ is a chain. In fact $\impossibility \snarrower \stmt_j \snarrower \stmt$ for all $j$ and $\stmt_i \narrower \stmt_j$ if $i \leq j$.

To show that $C$ is saturated, take a statement $\hat{\stmt} \narrower \stmt$ such that $\hat{\stmt} \notin C$. Let $A \subseteq X'$ be the set of possibilities compatible with $\hat{\stmt}$. If $A$ is infinite, since $\hat{\stmt} \nequiv \stmt$, there exists an $x_i \in X'$ such that $x_i \notin A$. Therefore $\stmt_{i} \nnarrower \hat{\stmt}$ since the first is compatible with a possibility that is not compatible with the second, and $\hat{\stmt} \nnarrower \stmt_{i}$ since $\hat{\stmt}$ is compatible with infinitely many statements and $\stmt_{i}$. If $A$ is finite, let the $x_i \in A$ for which $i$ is highest. Note that $A$ cannot contain all previous $x_j$ in the sequence, since $\hat{\stmt} \notin C$. Let $x_j$ be the possibility with the highest $j \leq i$ that is not in $A$. We have $\hat{\stmt} \nnarrower \stmt_{j}$ and $\stmt_{i} \nnarrower \hat{\stmt}$ since each is compatible with a possibility incompatible with the other. Since we cannot add other statements, $C$ is saturated.

Now we show that, even though $C$ is saturated, it does not go through all levels of granularity. Let $x \in X'$ and consider $\stmt \AND \NOT x$. This is finer than $\stmt$ and is coarser than all other statements in $C$ since they are all compatible with finitely many possibilities. Therefore SCP does not hold. Therefore, if SCP holds, the domain cannot have infinitely many possibilities.
\end{proof}
